Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2011-11-04 appendix.tex
ecology letters copy submitted

%        File: pnas_appendix.tex
%     Created: Wed Oct 12 05:00 PM 2011 P
% Last Change: Wed Oct 12 05:00 PM 2011 P
%

@@

\journal{Proceedings of the National Academy of Sciences } 
\begin{document}
%\begin{frontmatter}
%  \title{  }
%  \author[cpb]{Carl Boettiger\corref{cor1}}
%  \author[alan]{Alan Hastings}
%  \ead{cboettig@ucdavis.edu}
%  \cortext[cor1]{Corresponding author.}
%  \address[cpb]{Center for Population Biology, University of California, Davis, United States}
%  \address[esp]{Department of Environmental Science and Policy, University of California, Davis, United States}
%\end{frontmatter}
%
\appendix
\renewcommand*\thefigure{S\arabic{figure}}
\renewcommand*\theequation{S\arabic{equation}}
\section{Comparing Models}\label{Cox}
Likelihood methods form the basis of much of modern statistics, in both Frequentist and Bayesian paradigms.  
The ability evaluate likelihoods directly by computation has made it possible to treat cases that do not conform to traditional assumptions more directly.
The basis of likelihood comparisons has its roots in the Neyman-Pearson Lemma, 
which essentially asserts that comparing likelihoods is the most powerful test
of a choice between two hypotheses~\citep{Neyman1933}, which motivates
tests from the simple likelihood ratio test up through modern model adequacy methods.

@@

 \section{Distributions of the indicator statistic}
The ROC curves are generated from distributions of the measure of increase, $\tau$, in a warning statistic (variance, autocorrelation, etc.) or the likelihood ratio $\delta$ from the likelihood approach.  The value of the statistic observed in the original data is indicated by a black triangle. Deciding whether or not this observed value indicates an early warning or not depends on the location of the threshold.  Thresholds involving larger values of $\tau$ or $\delta$ will generally involve fewer false alarms, but also catch fewer of the true positives.  The large overlap between the distribution under the stable model (blue) and the deteriorating stability system (red) indicates how difficult it can be for such statistics to descriminate between these alternatives.  The nature of this trade-off is illustrated by the ROC curve, which can therefore be helpful in selecting the threshold for a particular statistic.

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS1.pdf}
  \end{center}
  \caption{Distribution of the indicator statistics corresponding to the ROC curves in Fig. 3, main text. As in Fig. 1, the expected distribution of the statistic under the null hypothesis of a stable system is shown in blue, while the distribution under the estimated deteriorating stability system is shown in red.  The value of the test statistic ($\delta$ in Likelihood, top row, otherwise $\tau$), observed in the original timeseries is indicated by the black triangle. Distributions are produced from a Gaussian kernel density estimate from 500 replicates, as described in Methods.}
  \label{fig:S1}
\end{figure}

>>>>

@@

\journal{a Letter to \emph{Ecology Letters}, Supplementary Information} 
\begin{document}
%\begin{frontmatter}
%  \title{  }
%  \author[cpb]{Carl Boettiger\corref{cor1}}
%  \author[alan]{Alan Hastings}
%  \ead{cboettig@ucdavis.edu}
%  \cortext[cor1]{Corresponding author.}
%  \address[cpb]{Center for Population Biology, University of California, Davis, United States}
%  \address[esp]{Department of Environmental Science and Policy, University of California, Davis, United States}
%\end{frontmatter}
%
\appendix
\renewcommand*\thefigure{S\arabic{figure}}
\renewcommand*\theequation{S\arabic{equation}}
\section{Comparing Models}\label{Cox}
Likelihood methods form the basis of much of modern statistics, in both Frequentist and Bayesian paradigms.  
The ability evaluate likelihoods directly by computation has made it possible to treat cases that do not conform to traditional assumptions more directly.
The basis of likelihood comparisons has its roots in the Neyman-Pearson Lemma, 
which essentially asserts that comparing likelihoods is the most powerful test
of a choice between two hypotheses~\citep{Neyman1933}, which motivates
tests from the simple likelihood ratio test up through modern model adequacy methods.

@@

 \section{Distributions of the indicator statistic}
The ROC curves are generated from distributions of the measure of increase, $\tau$, in a warning statistic (variance, autocorrelation, etc.) or the likelihood ratio $\delta$ from the likelihood approach.  The value of the statistic observed in the original data is indicated by a black triangle. Deciding whether or not this observed value indicates an early warning or not depends on the location of the threshold.  Thresholds involving larger values of $\tau$ or $\delta$ will generally involve fewer false alarms, but also catch fewer of the true positives.  The large overlap between the distribution under the stable model (blue) and the deteriorating stability system (red) indicates how difficult it can be for such statistics to discriminate between these alternatives.  The nature of this trade-off is illustrated by the ROC curve, which can therefore be helpful in selecting the threshold for a particular statistic.

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS1.pdf}
  \end{center}
  \caption{Distribution of the indicator statistics corresponding to the ROC curves in Fig. 3, main text. As in Fig. 1, the expected distribution of the statistic under the null hypothesis of a stable system is shown in blue, while the distribution under the estimated deteriorating stability system is shown in red.  The value of the test statistic ($\delta$ in Likelihood, top row, otherwise $\tau$), observed in the original time-series is indicated by the black triangle. Distributions are produced from a Gaussian kernel density estimate from 500 replicates, as described in Methods.}
  \label{fig:S1}
\end{figure}
